Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-07-22 threshold/draft.Rnw
Fixed column scaling in tcor.R

Let $A=U\Sigma V^T$ be the singular value decomposition of $A$, where
$U\in\mathcal{R}^{m\times k}$, $U^TU = I$, $V\in\mathcal{R}^{n\times k}$,
$V^TV=I$, and $\Sigma\in\mathcal{R}^{k\times k}$ is a diagonal matrix with
diagonal entries $\sigma_1 \ge \sigma_2 \ge \cdots \ge \sigma_k > 0$.  For
later convenience, we write $s_{1:p}=[\sigma_1, \sigma_2, \ldots, \sigma_p]^T$ to be
the vector of the first $p\le k$ singular values along the diagonal of
$\Sigma$.  Denote the columns of $V$ as $V=[v_1, v_2, \ldots, v_k]$, where each
$v_j\in\mathcal{R}^n$, $j=1,2,\ldots,k$.  Then $\mathrm{cor}(A) = A^TA =
V\Sigma^2 V^T$ is a symmetric eigenvalue decomposition of the correlation
matrix with eigenvalues $\sigma_1^2, \sigma_2^2, \ldots, \sigma_k^2$. The span
of the columns of $V$ form an orthonormal basis of the range of the correlation
matrix $A^TA$.

@@

Let $t$ be a given correlation threshold $0<t<1$, and let
$P\in\mathcal{R}^{n\times n}$ be a permutation that orders then entries of
$v_1$ in increasing order.  It follows from Equations \ref{cordist} and
\ref{vdist} that for any $i,j=1,2,\ldots,n$, if $\sigma_1^2 |v_{1i} - v_{1j}|^2
\ge 2(1-t)$ then the column vectors $a_j$ and $a_i$ do not meet the correlation
threshold.  Let $\ell$ be the longest run of successive points in $P v_1$
within the interval $\sqrt{2(1-t)}/\sigma_1$. For example, Figure \ref{fig1}
displays the entries of $v_1$ for the ``BicatYeast'' example from Section
\ref{examples} in increasing order.  The lines in the plot illustrate the
interval associated with the correlation threshold $t=0.99$. The quantity
$\ell$ is obtained by rolling the interval across all the points in the plot
and counting the maximum number of points that lie within the interval.  With
respect to the ordering $P$, $\ell$ represents the biggest index difference
that paris of indices corresponding to correlated vectors above the threshold
can exhibit.

@@

Let $D^i\in\mathcal{R}^{n-i, n}$, $i=1,2,\ldots,n-1$, be the $i$th order finite
difference matrix with $-1$ along the diagonal, $1$ along the $i$th
super-diagonal, and zeros elsewhere.  Consider just adjacent points of $P v_1$.
Then $D^1P v_1$ consists of the differences of adjacent entries in $P v_1$.
Entries such that $\sigma_1^2 (D^1 P v_1)^2 > 2(1-t)$ correspond to pairs of
vectors that do not meet the correlation threshold, where here and below
exponentiation applied to a vector denotes element-wise exponentiation.
However, this observation is unlikely to rule out many pairs of vectors in most
problems. For example, the minimum distance between the points shown in the
example in Figure \ref{fig1} is less that $\sqrt{2(1-t)}/\sigma_1$, and no
pruning at all occurs.

Including more terms from Equation \ref{vdist} increases our ability to prune
pairs of vectors below the correlation threshold. For example, including
only one more term finds that 72\% of adjacent vectors fall below the
correlation threshold
in the example shown in Figure \ref{fig1} and are pruned by identifying
indices such that
\[
\sigma_1^2 (D^1 P v_1)^2 +
\sigma_2^2 (D^1 P v_2)^2 > 2(1-t).
\]
Note that we use the permutation $P$ defined by the order of vector $v_1$
throughout. Including five terms prunes over 97\% of adjacent vector pairs,
leaving only 11 candidate pairs of adjacent vectors.
In general, using $p\le k$ terms to prune pairs of vectors below the
correlation threshold boils down to evaluating the expression
\begin{equation}\label{proj}
(D^1 P V_{1:p} )^2 s_{1:p}^2 > 2(1-t),
\end{equation}
where 
$V_{1:p}$ denotes the first $p$ columns of the
matrix $V$, $s_{1:p}$ the vector of the first $p$ singular values.

Pairs of vectors below the correlation threshold that lie more than one index
difference apart relative to the permutation $P$ can be similarly pruned by
replacing $D^1$ with $D^j$, $j=1,2,\ldots,\ell$ in the above expressions.
Following this procedure, we can produce a well-pruned candidate set of
possible vector pairs that meet a given correlation threshold using $\ell$
matrix vector products with matrices of order $n \times p$, where $p$ is chosen
in practice such that $p<<k$. The steps of this procedure are outlined in the
following algorithm:

\begin{algorithm}\label{prune}
Input: data matrix $A\in \mathcal{R}^{m\times n}$, correlation threshold $t$, maximum
truncated SVD rank $p<<k$.
\begin{enumerate}
\item Compute the truncated SVD $AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$,
where $V_{1:p} = [v_1, v_2, \ldots, v_p]$
and $\mathrm{diag}(s_{1:p})$ denotes the diagonal matrix of the first $p$ singular values.
\item Compute permutation $P$ that orders points in $v_1$.
\item Compute $\ell$, the longest run of successive points in $P v_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$.
\item Compute a set of candidate pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
\[
\bigcup_{\tilde{i}=1}^\ell
\left\{
(\tilde{j},\tilde{j}+\tilde{i}) : 
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
\right\},
\]
where $e_{\tilde{j}}\in\mathcal{R}^n$ consists of $1$ in position $\tilde{j}$ and
zeros otherwise.
\item If there are judged to be  ``too many'' candidate pairs, increase $p$ and go to
step 1. Continue to increase $p$ in this way until the number of candidate
pairs stops decreasing (the best this algorithm can do).
\item Recover the non-permuted candidate column vector
indices $(j,i)$ from $j=e_{\tilde{j}}^T Pq$ and $i=e_{\tilde{j}+\tilde{i}}^TPq$ for every
pair $(\tilde{j},\tilde{j} + \tilde{i})$ in step 4, where
$q\in\mathcal{R}^n = [1,2,\ldots,n]^T$.
\item Compute the full correlation coefficient for each column vector pair identified in
the previous step, applying the correlation threshold to this reduced set of values.
\end{enumerate}
\end{algorithm}
Because the singular values $\sigma_j$, $j=1,2,\ldots,n$ are non-increasing and
decrease in proportion to variability in the data, it's often reasonable to
judge whether there are ``too many'' candidate pairs (step 5) after only a 1-d
projection $\tilde{i}=1$ in step 4 of the algorithm.

@@

The augmented, implicitly restarted Lanczos bidiagonalization algorithm (IRLBA)
of Baglama and Reichel \cite{br} efficiently computes truncated singular value
decompositions of large dense or sparse matrices. Reference implementations are
available for R\cite{irlba-R}, Matlab\cite{irlba-M}, and Python\cite{irlba-P}.
The R implementation includes efficient methods for re-starting the algorithm
to increase the subspace dimension (see step 5 of Algorithm \ref{prune}), and
efficient methods for computing the SVD of scaled and centered matrices without
explicitly forming a new matrix.

>>>>

Let $A=U\Sigma V^T$ be the singular value decomposition of $A$, where
$U\in\mathcal{R}^{m\times k}$, $U^TU = I$, $V\in\mathcal{R}^{n\times k}$,
$V^TV=I$, and $\Sigma\in\mathcal{R}^{k\times k}$ is a diagonal matrix with
diagonal entries $\sigma_1 \ge \sigma_2 \ge \cdots \ge \sigma_k > 0$.  For
later convenience, we write $s_{1:p}=[\sigma_1, \sigma_2, \ldots, \sigma_p]^T$
to be the vector of the first $p\le k$ singular values along the diagonal of
$\Sigma$.  Denote the columns of $V$ as $V=[v_1, v_2, \ldots, v_k]$, where each
$v_j\in\mathcal{R}^n$, $j=1,2,\ldots,k$.  Then $\mathrm{cor}(A) = A^TA =
V\Sigma^2 V^T$ is a symmetric eigenvalue decomposition of the correlation
matrix with nonzero eigenvalues $\sigma_1^2, \sigma_2^2, \ldots, \sigma_k^2$.
The span of the columns of $V$ form an orthonormal basis of the range of the
correlation matrix $A^TA$.

@@

Let $t$ be a given correlation threshold $0<t<1$, and let
$P\in\mathcal{R}^{n\times n}$ be a permutation matrix that orders then entries
of $v_1$ in increasing order.  It follows from Equations \ref{cordist} and
\ref{vdist} that for any $i,j=1,2,\ldots,n$, if $\sigma_1^2 |v_{1i} - v_{1j}|^2
\ge 2(1-t)$ then the column vectors $a_j$ and $a_i$ do not meet the correlation
threshold.  Let $\ell$ be the longest run of successive points in $P v_1$
within the interval $\sqrt{2(1-t)}/\sigma_1$. For example, Figure \ref{fig1}
displays the entries of $v_1$ for the ``BicatYeast'' example from Section
\ref{examples} in increasing order.  The lines in the plot illustrate the
interval associated with the correlation threshold $t=0.99$. The quantity
$\ell$ is obtained by rolling the interval across all the points in the plot
and counting the maximum number of points that lie within the interval.  With
respect to the ordering $P$, $\ell$ represents the biggest index difference
that paris of indices corresponding to correlated vectors above the threshold
can exhibit.

@@

Let $D^i\in\mathcal{R}^{n-i, n}$, $i=1,2,\ldots,n-1$, be the $i$th order finite
difference matrix with $-1$ along the diagonal, $1$ along the $i$th
super-diagonal, and zeros elsewhere.  Consider just adjacent points of $P v_1$.
Then $D^1P v_1$ consists of the differences of adjacent entries in $P v_1$.
Entries such that $\sigma_1^2 (D^1 P v_1)^2 > 2(1-t)$ correspond to pairs of
vectors that do not meet the correlation threshold, where here and below
an exponent applied to a vector denotes element-wise exponentiation.
However, this observation is unlikely to rule out many pairs of vectors in most
problems. For example, the minimum distance between the points shown in the
example in Figure \ref{fig1} is less that $\sqrt{2(1-t)}/\sigma_1$, and no
pruning at all occurs.

Including more terms from Equation \ref{vdist} increases our ability to prune
pairs of vectors below the correlation threshold. For example, including only
one more term finds that 72\% of adjacent vectors fall below the a correlation
threshold of $0.99$ in the example shown in Figure \ref{fig1} and are pruned by
identifying indices such that
\[
\sigma_1^2 (D^1 P v_1)^2 +
\sigma_2^2 (D^1 P v_2)^2 > 2(1-t).
\]
Note that we use the permutation $P$ defined by the order of vector $v_1$
throughout. Including five terms prunes over 97\% of adjacent vector pairs,
leaving only 11 candidate pairs of adjacent vectors that might possibly
meet the correlation threshold.
In general, using $p\le k$ terms to prune pairs of vectors below the
correlation threshold boils down to evaluating the expression
\begin{equation}\label{proj}
(D^1 P V_{1:p} )^2 s_{1:p}^2 > 2(1-t),
\end{equation}
where 
$V_{1:p}$ denotes the first $p$ columns of the
matrix $V$, $s_{1:p}$ the vector of the first $p$ singular values.

Pairs of vectors below the correlation threshold that lie more than one index
difference apart relative to the permutation $P$ can be similarly pruned by
replacing $D^1$ with $D^j$, $j=1,2,\ldots,\ell$ in the above expressions.
Following this process, we can produce a well-pruned candidate set of possible
vector pairs that meet a given correlation threshold using $\ell$ matrix vector
products with matrices of order $n \times p$, where $p$ is chosen in practice
such that $p<<k$. These steps are formalized in Algorithm \ref{prune} below.

\begin{algorithm}\label{prune}
Input: data matrix $A\in \mathcal{R}^{m\times n}$, correlation threshold $t$,
truncated SVD rank $p<<k$.
\begin{enumerate}
\item Compute the truncated SVD $AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$,
where $V_{1:p} = [v_1, v_2, \ldots, v_p]$
and $\mathrm{diag}(s_{1:p})$ denotes the diagonal matrix of the first $p$ singular values.
\item Compute permutation $P$ that orders points in $v_1$.
\item Compute $\ell$, the longest run of successive points in $P v_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$.
\item Compute a set of candidate pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
\[
\bigcup_{\tilde{i}=1}^\ell
\left\{
(\tilde{j},\tilde{j}+\tilde{i}) : 
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
\right\},
\]
where $e_{\tilde{j}}\in\mathcal{R}^n$ consists of $1$ in position $\tilde{j}$ and
zeros otherwise.
\item If there are judged to be  ``too many'' candidate pairs, increase $p$ and go to
step 1. Continue to increase $p$ in this way until the number of candidate
pairs stops decreasing (the best this algorithm can do).
\item Recover the non-permuted candidate column vector
indices $(j,i)$ from $j=e_{\tilde{j}}^T Pq$ and $i=e_{\tilde{j}+\tilde{i}}^TPq$ for every
pair $(\tilde{j},\tilde{j} + \tilde{i})$ in step 4, where
$q\in\mathcal{R}^n = [1,2,\ldots,n]^T$.
\item Compute the full correlation coefficient for each column vector pair identified in
the previous step, applying the correlation threshold to this reduced set of values.
\end{enumerate}
\end{algorithm}
Because the singular values $\sigma_j$, $j=1,2,\ldots,n$ are non-increasing and
decrease in proportion to variability in the data, it's often reasonable to
judge whether there are ``too many'' candidate pairs (step 5) after only a 1-d
projection $\tilde{i}=1$ in step 4 of the algorithm.

@@

The augmented implicitly restarted Lanczos bidiagonalization algorithm (IRLBA)
of Baglama and Reichel \cite{br} efficiently computes truncated singular value
decompositions of large dense or sparse matrices. Reference implementations are
available for R\cite{irlba-R}, Matlab\cite{irlba-M}, and Python\cite{irlba-P}.
The R implementation includes efficient methods for re-starting the algorithm
to increase the subspace dimension (see step 5 of Algorithm \ref{prune}), and
efficient methods for computing the SVD of scaled and centered matrices without
explicitly forming a new matrix.

Let $x\in\mathcal{R}^n$ represent the vector of column means of the matrix $A$,
$W\in\mathcal{R}^{n\times n}$ be a diagonal matrix with the reciprocal of the
column norms of $A$ along the diagonal, and let $e\in\mathcal{R}^m$ be a vector
of all ones. Then $(A - ex^T)W$ is a centered matrix with zero column means,
and scaled to have unit column norms, and

XXX

IRLBA, based on the Lanczos process, is an iterative method based on
matrix-vector products involving the matrix $A$.
